---
title: "020-data-source-registry"
date: today
params:
  name: "020-data-source-registry"
bibliography: references.bib
---

**Updated: `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'CET')` CET.**

Registers the public data sources backing the pipeline. Each source is one
row in `data_source`. Raw supplementary tables live under
`data/00-raw/d020-data-source-registry/<source-key>/`.

**Ripple-1 scope** (`020-r1`) — seven sources: all six Geng et al. modeling
bacteriome cohorts plus the Yahara virome.

- `geng2024_<bioproject>` (×6) — periodontitis bacteriome cohorts from the
  Geng et al. 2024 meta-analysis [@2024-Universal_Geng]; cohort-level facts
  (n.case / n.control, country, instrument) are pulled from sheet `Table S1`
  of the supplementary table.
- `yahara2021` — long-read oral phageome from PromethION saliva sequencing
  [@2021-Longread_Yahara]. 4 saliva samples (one healthy donor, four
  conditions). VIRSorter-derived phage contigs.

Ripple-2 (`020-r2`) will pair each Geng cohort with a body-site- and
disease-matched virome dataset.

```{r}
#| label: params
#| eval: !expr interactive()
#| include: false
params <- list(name = "020-data-source-registry")
```

```{r}
#| label: setup
#| include: false
#| message: false
#| warning: false
#| cache: false
wd <- "analyses"
if (basename(getwd()) != wd) {
    setwd(here::here(wd))
}
here::i_am(paste0(params$name, ".qmd"), uuid = "fa33ff54-384c-4e3f-be14-cf6aa3347ecb")
qproj::proj_create_dir_target(params$name, clean = FALSE)

path_target <- qproj::proj_path_target(params$name)
path_source <- qproj::proj_path_source(params$name)
path_raw <- path_source("00-raw")
path_resource <- here::here(path_raw, "d00-resource")
path_data <- here::here(path_raw, paste0("d", params$name))
dir.create(path_raw, recursive = TRUE, showWarnings = FALSE)
dir.create(path_data, recursive = TRUE, showWarnings = FALSE)
dir.create(path_resource, recursive = TRUE, showWarnings = FALSE)
```

```{r}
#| label: packages
#| message: false
#| warning: false
#| cache: false
suppressPackageStartupMessages({
  library(here)
  library(conflicted)
  library(tidyverse)
  library(readxl)
  library(jsonlite)
  library(DBI)
  devtools::load_all()
})
conflicted::conflicts_prefer(
  dplyr::filter,
  dplyr::lag,
  dplyr::first,
  dplyr::last,
  dplyr::between,
  .quiet = TRUE
)
```

## Verify raw inputs are present

The supplementary tables were fetched into
`data/00-raw/d020-data-source-registry/`. The render fails fast if anything
is missing.

```{r}
#| label: verify-raw
path_geng_repo <- here::here(path_data, "wei2024-imeta-repo")
path_geng_supp <- here::here(path_geng_repo, "Supplementary table",
                             "Supplementary Table.xlsx")
path_geng_rdata <- here::here(path_geng_repo, "Rdata", "Figure 01.Rdata")
path_yahara_dir <- here::here(path_data, "yahara2021")
path_yahara_supp4 <- here::here(path_yahara_dir, "MOESM7.xlsx")  # Supp Data 4 = viral seqs
path_yahara_supp5 <- here::here(path_yahara_dir, "MOESM8.xlsx")  # Supp Data 5 = Strep phages

stopifnot(
  file.exists(path_geng_supp),
  file.exists(path_geng_rdata),
  file.exists(path_yahara_supp4),
  file.exists(path_yahara_supp5)
)

df_inputs <- tibble::tibble(
  source = c("geng2024", "geng2024", "yahara2021", "yahara2021"),
  file   = c(basename(path_geng_supp), basename(path_geng_rdata),
             basename(path_yahara_supp4), basename(path_yahara_supp5)),
  bytes  = c(file.size(path_geng_supp), file.size(path_geng_rdata),
             file.size(path_yahara_supp4), file.size(path_yahara_supp5))
)
knitr::kable(df_inputs, caption = "Raw supplementary inputs (verified present)")
```

## Build the registry rows

The Geng cohorts are read from `Table S1` of the supplementary spreadsheet.
We keep only the six modeling cohorts (the two validation cohorts are not
shipped in the `Figure 01.Rdata` feat_list and are out of ripple-1 scope).

```{r}
#| label: build-geng-rows
df_table_s1 <- readxl::read_excel(path_geng_supp, sheet = "Table S1") |>
  dplyr::filter(`modeling or validation` == "modeling")

geng_table_url <- "https://github.com/whchenlab/Oral-Microbiome-Based-Signature-for-Periodontitis"

df_geng <- df_table_s1 |>
  dplyr::transmute(
    source_id        = paste0("geng2024_", project_id),
    citation_key     = "2024-Universal_Geng",
    doi              = "10.1002/imt2.212",
    source_kind      = "bacteriome",
    n_samples        = as.integer(num.case + num.control),
    table_url        = geng_table_url,
    notes            = sprintf(
      "Ripple-1 M1a; %s saliva/oral cohort %s. %d case / %d control. %s.",
      country_fullname, project_id, num.case, num.control, instrument_model
    )
  )

df_yahara <- tibble::tibble(
  source_id    = "yahara2021",
  citation_key = "2021-Longread_Yahara",
  doi          = "10.1038/s41467-020-20199-9",
  source_kind  = "virome",
  n_samples    = 4L,
  table_url    = "https://www.nature.com/articles/s41467-020-20199-9#Sec19",
  notes        = "MVP M2a; long-read PromethION saliva phageome. 1 healthy donor x 4 conditions. VIRSorter-derived contigs."
)

df_sources <- dplyr::bind_rows(df_geng, df_yahara)
df_sources
```

## Write to the contract layer

```{r}
#| label: write-db
con <- load_db()
assert_schema(con)
write_table_db(con, "data_source", df_sources, append = FALSE)
df_after <- read_table_db(con, "data_source")
close_db(con)
knitr::kable(df_after, caption = "data_source after registry write")
```

## Persist the registry table for human reference

```{r}
#| label: persist
write_csv(df_sources, path_target("data_source.csv"))
```

## Files written

These files have been written to the target directory, `r paste0("data/", params$name)`:

```{r}
#| label: list-files-target
knitr::kable(qproj::proj_dir_info(path_target(), tz = "CET"))
```
