---
title: "030-preprocessing-pipeline"
date: today
params:
  name: "030-preprocessing-pipeline"
---

**Updated: `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'CET')` CET.**

::: {.callout-note}
## Deferred to ripple

This module is intentionally a stub in the MVP. The MVP loads
**published abundance and vOTU tables** directly (modules 040 / 070), so no
fastq → species pipeline is required to validate the contract layer.

**Ripple plan** (`030-r1` / `030-r2`):

- `030-r1`: Nextflow pipeline on SLURM that re-processes one published cohort
  (`geng2024_PRJDB11203` SRA fastq) with vpipe / Kraken2 / MetaPhlAn4 →
  species table. Validate by comparing species relative abundances against the
  published `Figure 01.Rdata` matrix; flag any species with > 25% disagreement.
- `030-r2`: extend to virome on SLURM (ViWrap / ViOTUcluster / CheckV) and
  reproduce the Yahara 2021 vOTU catalog from raw long reads.

See `IMPLEMENTATION_PLAN.md` and `/Users/cmbjx/.claude/plans/mvp-linear-ripple.md`
§Ripple roadmap for the full ripple ladder.
:::

```{r}
#| label: params
#| eval: !expr interactive()
#| include: false
params <- list(name = "030-preprocessing-pipeline")
```

```{r}
#| label: setup
#| include: false
#| message: false
#| warning: false
#| cache: false
wd <- "analyses"
if (basename(getwd()) != wd) {
    setwd(here::here(wd))
}
here::i_am(paste0(params$name, ".qmd"), uuid = "dac3bb52-6118-454c-8999-b539b683d397")
qproj::proj_create_dir_target(params$name, clean = FALSE)

path_target <- qproj::proj_path_target(params$name)
path_source <- qproj::proj_path_source(params$name)
```

## Status

```{r}
#| label: status
tibble::tibble(
  module = "030-preprocessing-pipeline",
  status = "deferred-to-ripple",
  ripple_targets = "030-r1 (bacteriome fastq), 030-r2 (virome fastq)",
  next_action = "After MVP DoD passes, scaffold a Nextflow pipeline under nextflow/"
) |> knitr::kable(caption = "Module status")
```

## Files written

These files have been written to the target directory, `r paste0("data/", params$name)`:

```{r}
#| label: list-files-target
knitr::kable(qproj::proj_dir_info(path_target(), tz = "CET"))
```
