---
title: "050-m1a-differential-analysis"
date: today
params:
  name: "050-m1a-differential-analysis"
bibliography: references.bib
---

**Updated: `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'CET')` CET.**

Multi-method, single-cohort differential abundance analysis on the
Geng PRJDB11203 saliva sub-cohort. The MVP baseline (Wilcoxon rank-sum +
Benjamini–Hochberg FDR on TSS-normalized relative abundances) is preserved
verbatim. Ripple `050-r1` adds three additional methods (MaAsLin2, ANCOMBC,
DESeq2) and emits a consensus candidate set requiring agreement (q < 0.1 +
consistent direction) in ≥ 2 of the 4 methods. Output goes into
`candidate_microbe` (phenotype = "periodontitis").

**Ripple ladder**:

- `050-r1` (this revision): MaAsLin2 + ANCOMBC + DESeq2 alongside Wilcoxon,
  with consensus requiring q < 0.1 in ≥ 2 of 4 methods + matching direction.
- `050-r2`: cross-cohort meta-analysis (random-effects) over the seven Geng
  cohorts; emit candidates with `n_cohorts ≥ 3`.
- `050-r3`: pathway-level candidates from HUMAnN3 in addition to species.

```{r}
#| label: params
#| eval: !expr interactive()
#| include: false
params <- list(name = "050-m1a-differential-analysis")
```

```{r}
#| label: setup
#| include: false
#| message: false
#| warning: false
#| cache: false
wd <- "analyses"
if (basename(getwd()) != wd) {
    setwd(here::here(wd))
}
here::i_am(paste0(params$name, ".qmd"), uuid = "07ad2b48-bbcc-4ba2-86f1-d4ad7c7ec6cd")
qproj::proj_create_dir_target(params$name, clean = FALSE)

path_target <- qproj::proj_path_target(params$name)
```

```{r}
#| label: packages
#| message: false
#| warning: false
#| cache: false
suppressPackageStartupMessages({
  library(here)
  library(conflicted)
  library(tidyverse)
  library(digest)
  library(DBI)
  library(Maaslin2)
  library(ANCOMBC)
  library(DESeq2)
  library(phyloseq)
  devtools::load_all()
})
conflicted::conflicts_prefer(
  dplyr::filter,
  dplyr::lag,
  dplyr::first,
  dplyr::last,
  dplyr::between,
  dplyr::desc,
  dplyr::collapse,
  dplyr::count,
  dplyr::rename,
  dplyr::select,
  dplyr::slice,
  dplyr::combine,
  base::setdiff,
  base::intersect,
  base::union,
  base::Reduce,
  base::Position,
  base::Find,
  base::Filter,
  base::Map,
  base::strsplit,
  stats::IQR,
  base::which,
  .quiet = TRUE
)
```

## Pull samples + abundances from the contract layer

```{r}
#| label: pull
con <- load_db()
df_sample <- read_table_db(con, "sample") |>
  dplyr::filter(source_id == "geng2024_PRJDB11203")
df_profile <- read_table_db(con, "taxon_profile") |>
  dplyr::filter(sample_id %in% df_sample$sample_id, taxon_kind == "bacterium")
close_db(con)

df_wide <- df_profile |>
  dplyr::select(sample_id, taxon_id, taxon_name, abundance) |>
  tidyr::pivot_wider(id_cols = c(taxon_id, taxon_name),
                     names_from = sample_id, values_from = abundance,
                     values_fill = 0)

cat("Wide matrix:", nrow(df_wide), "species x",
    ncol(df_wide) - 2, "samples\n")
```

## Wilcoxon rank-sum (peri vs healthy)

```{r}
#| label: wilcoxon
ids_case <- df_sample$sample_id[df_sample$disease_status == "periodontitis"]
ids_ctrl <- df_sample$sample_id[df_sample$disease_status == "healthy"]
stopifnot(length(ids_case) >= 5, length(ids_ctrl) >= 5)

# species must be present (>0) in >= 10% of all samples to be tested
mat_abund <- as.matrix(df_wide[, c(ids_case, ids_ctrl)])
rownames(mat_abund) <- df_wide$taxon_id
prevalence <- rowMeans(mat_abund > 0)
keep <- prevalence >= 0.10

df_test <- tibble::tibble(
  taxon_id   = df_wide$taxon_id[keep],
  taxon_name = df_wide$taxon_name[keep]
) |>
  dplyr::mutate(
    mean_case = rowMeans(mat_abund[keep, ids_case, drop = FALSE]),
    mean_ctrl = rowMeans(mat_abund[keep, ids_ctrl, drop = FALSE]),
    p_value   = purrr::map2_dbl(
      asplit(mat_abund[keep, ids_case, drop = FALSE], 1),
      asplit(mat_abund[keep, ids_ctrl, drop = FALSE], 1),
      function(a, b) {
        suppressWarnings(stats::wilcox.test(a, b, exact = FALSE)$p.value)
      }
    ),
    effect_size = log2((mean_case + 1e-6) / (mean_ctrl + 1e-6)),
    direction   = ifelse(effect_size > 0, "disease_enriched", "health_enriched"),
    q_value     = stats::p.adjust(p_value, method = "BH")
  )
cat("Tested species:", nrow(df_test), "| q<0.1:", sum(df_test$q_value < 0.1, na.rm = TRUE), "\n")
```

## Build Wilcoxon `candidate_microbe` rows (MVP baseline, preserved)

We keep all tested species with `q_value < 0.1` plus, if fewer than 30 pass,
top-up with the smallest-p species so the MVP downstream has signal. (Ripple
will tighten this back to a strict q<0.1 once meta-analysis is in place.)

```{r}
#| label: build-candidates
df_pass <- df_test |>
  dplyr::arrange(p_value) |>
  dplyr::filter(q_value < 0.1)

if (nrow(df_pass) < 30) {
  df_topup <- df_test |>
    dplyr::arrange(p_value) |>
    dplyr::filter(!taxon_id %in% df_pass$taxon_id) |>
    head(30 - nrow(df_pass))
  df_candidates <- dplyr::bind_rows(df_pass, df_topup)
} else {
  df_candidates <- df_pass
}

ver <- schema_version()

df_candidate_microbe <- df_candidates |>
  dplyr::mutate(
    phenotype      = "periodontitis",
    n_cohorts      = 1L,
    method         = "wilcoxon",
    schema_version = ver,
    candidate_id   = paste0(
      "CAND_",
      vapply(paste(taxon_id, phenotype, method, sep = "|"),
             function(s) substr(digest::digest(s, algo = "xxhash64"), 1, 12),
             character(1))
    )
  ) |>
  dplyr::select(candidate_id, taxon_id, taxon_name, phenotype, direction,
                effect_size, p_value, q_value, n_cohorts, method, schema_version)

stopifnot(!any(duplicated(df_candidate_microbe$candidate_id)))
df_candidate_microbe |>
  head(20) |>
  knitr::kable(caption = "Top 20 candidate microbes (smallest p)")
```

```{r}
#| label: direction-balance
dplyr::count(df_candidate_microbe, direction) |>
  knitr::kable(caption = "Direction balance among Wilcoxon candidates")
```

## Helper: rows in the long candidate format

```{r}
#| label: helper-rows
make_candidate_rows <- function(df, method_name) {
  df |>
    dplyr::mutate(
      phenotype      = "periodontitis",
      n_cohorts      = 1L,
      method         = method_name,
      schema_version = ver,
      candidate_id   = paste0(
        "CAND_",
        vapply(paste(taxon_id, phenotype, method, sep = "|"),
               function(s) substr(digest::digest(s, algo = "xxhash64"), 1, 12),
               character(1))
      )
    ) |>
    dplyr::select(candidate_id, taxon_id, taxon_name, phenotype, direction,
                  effect_size, p_value, q_value, n_cohorts, method,
                  schema_version)
}

# Reused inputs for every method
mat_filt   <- mat_abund[keep, , drop = FALSE]                   # filtered abundance
df_features <- df_wide |> dplyr::filter(taxon_id %in% rownames(mat_filt)) |>
  dplyr::select(taxon_id, taxon_name)
df_meta    <- tibble::tibble(
  sample_id      = c(ids_case, ids_ctrl),
  disease_status = c(rep("periodontitis", length(ids_case)),
                     rep("healthy",       length(ids_ctrl)))
) |>
  dplyr::mutate(disease_status = factor(disease_status,
                                        levels = c("healthy", "periodontitis")))
df_meta_df <- as.data.frame(df_meta); rownames(df_meta_df) <- df_meta$sample_id
mat_for_methods <- mat_filt[, df_meta$sample_id, drop = FALSE]
```

## MaAsLin2

```{r}
#| label: maaslin2
#| message: false
#| warning: false
path_maaslin <- path_target("maaslin2_out")
if (dir.exists(path_maaslin)) unlink(path_maaslin, recursive = TRUE)

# Maaslin2 wants samples as rows, features as cols
mat_maaslin <- t(mat_for_methods)

fit_maaslin <- Maaslin2::Maaslin2(
  input_data        = as.data.frame(mat_maaslin),
  input_metadata    = df_meta_df,
  output            = path_maaslin,
  fixed_effects     = "disease_status",
  reference         = "disease_status,healthy",
  normalization     = "TSS",
  transform         = "LOG",
  analysis_method   = "LM",
  min_prevalence    = 0.10,
  min_abundance     = 0,
  max_significance  = 1,
  plot_heatmap      = FALSE,
  plot_scatter      = FALSE,
  standardize       = FALSE
)

df_maaslin <- tibble::as_tibble(fit_maaslin$results) |>
  dplyr::filter(metadata == "disease_status",
                value    == "periodontitis") |>
  dplyr::transmute(
    taxon_id    = feature,
    effect_size = coef,
    p_value     = pval,
    q_value     = qval,
    direction   = ifelse(coef > 0, "disease_enriched", "health_enriched")
  ) |>
  dplyr::left_join(df_features, by = "taxon_id") |>
  dplyr::filter(!is.na(p_value))

cat("MaAsLin2: tested", nrow(df_maaslin),
    "| q<0.1:", sum(df_maaslin$q_value < 0.1, na.rm = TRUE), "\n")

df_cand_maaslin <- df_maaslin |>
  dplyr::filter(q_value < 0.1) |>
  make_candidate_rows("maaslin2")
```

## ANCOMBC

```{r}
#| label: ancombc
#| message: false
#| warning: false
# ANCOMBC needs counts; round(rel_abund * 1e6) is the standard trick.
mat_counts <- round(mat_for_methods * 1e6)
storage.mode(mat_counts) <- "integer"

ps_otu  <- phyloseq::otu_table(mat_counts, taxa_are_rows = TRUE)
ps_samp <- phyloseq::sample_data(df_meta_df)
ps_obj  <- phyloseq::phyloseq(ps_otu, ps_samp)

fit_ancom <- ANCOMBC::ancombc(
  data         = ps_obj,
  formula      = "disease_status",
  p_adj_method = "BH",
  prv_cut      = 0.10,
  group        = "disease_status",
  struc_zero   = FALSE,
  neg_lb       = FALSE,
  conserve     = TRUE,
  verbose      = FALSE
)

df_ancom_raw <- tibble::tibble(
  taxon_id    = fit_ancom$res$lfc$taxon,
  effect_size = fit_ancom$res$lfc[["disease_statusperiodontitis"]],
  p_value     = fit_ancom$res$p_val[["disease_statusperiodontitis"]],
  q_value     = fit_ancom$res$q_val[["disease_statusperiodontitis"]]
) |>
  dplyr::mutate(
    direction = ifelse(effect_size > 0, "disease_enriched", "health_enriched")
  ) |>
  dplyr::left_join(df_features, by = "taxon_id") |>
  dplyr::filter(!is.na(p_value))

cat("ANCOMBC: tested", nrow(df_ancom_raw),
    "| q<0.1:", sum(df_ancom_raw$q_value < 0.1, na.rm = TRUE), "\n")

df_cand_ancom <- df_ancom_raw |>
  dplyr::filter(q_value < 0.1) |>
  make_candidate_rows("ancombc")
```

## DESeq2

```{r}
#| label: deseq2
#| message: false
#| warning: false
# Same count trick. DESeq2 requires integer counts.
dds <- DESeq2::DESeqDataSetFromMatrix(
  countData = mat_counts,
  colData   = df_meta_df,
  design    = ~ disease_status
)
# poscounts size factors handle zero-rich microbiome data better than RLE
dds <- DESeq2::estimateSizeFactors(dds, type = "poscounts")
dds <- DESeq2::DESeq(dds, quiet = TRUE)
res <- DESeq2::results(dds,
                      contrast = c("disease_status", "periodontitis", "healthy"))
df_deseq <- tibble::tibble(
  taxon_id    = rownames(res),
  effect_size = res$log2FoldChange,
  p_value     = res$pvalue,
  q_value     = stats::p.adjust(res$pvalue, method = "BH")
) |>
  dplyr::mutate(
    direction = ifelse(effect_size > 0, "disease_enriched", "health_enriched")
  ) |>
  dplyr::left_join(df_features, by = "taxon_id") |>
  dplyr::filter(!is.na(p_value))

cat("DESeq2: tested", nrow(df_deseq),
    "| q<0.1:", sum(df_deseq$q_value < 0.1, na.rm = TRUE), "\n")

df_cand_deseq <- df_deseq |>
  dplyr::filter(q_value < 0.1) |>
  make_candidate_rows("deseq2")
```

## Wilcoxon → method-tagged rows (full results)

```{r}
#| label: wilcoxon-full
df_wilcox_full <- df_test |>
  dplyr::transmute(taxon_id, taxon_name, effect_size, p_value, q_value,
                   direction)
```

## Method agreement diagnostics

```{r}
#| label: agreement
df_methods <- list(
  wilcoxon = df_wilcox_full,
  maaslin2 = df_maaslin,
  ancombc  = df_ancom_raw,
  deseq2   = df_deseq
)

# significant set per method (q<0.1, with direction)
sig_sets <- purrr::imap(df_methods, function(df, m) {
  df |>
    dplyr::filter(q_value < 0.1) |>
    dplyr::transmute(taxon_id, direction)
})

df_agree <- tidyr::expand_grid(a = names(sig_sets), b = names(sig_sets)) |>
  dplyr::filter(a < b) |>
  dplyr::mutate(
    n_a       = vapply(a, function(x) nrow(sig_sets[[x]]), integer(1)),
    n_b       = vapply(b, function(x) nrow(sig_sets[[x]]), integer(1)),
    n_overlap = purrr::map2_int(a, b, function(x, y) {
      length(intersect(sig_sets[[x]]$taxon_id, sig_sets[[y]]$taxon_id))
    }),
    jaccard = purrr::map2_dbl(a, b, function(x, y) {
      ix <- intersect(sig_sets[[x]]$taxon_id, sig_sets[[y]]$taxon_id)
      un <- union(sig_sets[[x]]$taxon_id, sig_sets[[y]]$taxon_id)
      if (length(un) == 0) NA_real_ else length(ix) / length(un)
    })
  )
knitr::kable(df_agree, digits = 3,
             caption = "Pairwise overlap among q<0.1 sets across methods")
```

## Consensus candidate set

A species enters the consensus set if it is significant (q<0.1) in
≥ 2 of the 4 methods AND all flagging methods agree on direction. The
consensus row's effect-size is the mean of the flagging methods' effect-sizes,
and the q-value is the **maximum** across flagging methods (conservative).

```{r}
#| label: consensus
df_long_sig <- purrr::imap_dfr(df_methods, function(df, m) {
  df |>
    dplyr::filter(q_value < 0.1) |>
    dplyr::mutate(method = m)
})

df_consensus <- df_long_sig |>
  dplyr::group_by(taxon_id, taxon_name) |>
  dplyr::summarise(
    n_methods         = dplyr::n(),
    n_dir_disease     = sum(direction == "disease_enriched"),
    n_dir_health      = sum(direction == "health_enriched"),
    effect_size       = mean(effect_size, na.rm = TRUE),
    q_value           = max(q_value,      na.rm = TRUE),
    methods_flagging  = paste(sort(method), collapse = ","),
    .groups = "drop"
  ) |>
  dplyr::filter(n_methods >= 2,
                n_dir_disease == 0 | n_dir_health == 0) |>
  dplyr::mutate(
    direction = ifelse(n_dir_disease > 0, "disease_enriched", "health_enriched"),
    p_value   = NA_real_
  )

cat("Consensus species (q<0.1 in >=2 methods + concordant direction):",
    nrow(df_consensus), "\n")

df_cand_consensus <- df_consensus |>
  dplyr::select(taxon_id, taxon_name, effect_size, p_value, q_value,
                direction) |>
  make_candidate_rows("consensus")

if (nrow(df_consensus) > 0) {
  df_consensus |>
    dplyr::arrange(q_value) |>
    head(15) |>
    dplyr::select(taxon_id, taxon_name, n_methods, direction,
                  effect_size, q_value, methods_flagging) |>
    knitr::kable(digits = 3,
                 caption = "Top 15 consensus species (sorted by q_value)")
}
```

## Combine all methods + write to the contract layer

```{r}
#| label: write-db
df_all_methods <- dplyr::bind_rows(
  df_candidate_microbe,    # wilcoxon (preserved MVP block)
  df_cand_maaslin,
  df_cand_ancom,
  df_cand_deseq,
  df_cand_consensus
)

stopifnot(!any(duplicated(df_all_methods$candidate_id)))

con <- load_db()
assert_schema(con)
DBI::dbExecute(con, "DELETE FROM candidate_microbe WHERE phenotype = 'periodontitis'")
DBI::dbWriteTable(con, "candidate_microbe", df_all_methods,
                  append = TRUE, row.names = FALSE)
df_summary <- DBI::dbGetQuery(
  con,
  "SELECT method, direction, COUNT(*) AS n
     FROM candidate_microbe
    WHERE phenotype = 'periodontitis'
 GROUP BY method, direction
 ORDER BY method, direction"
)
n_total <- DBI::dbGetQuery(con,
  "SELECT COUNT(*) AS n FROM candidate_microbe")$n
close_db(con)

knitr::kable(df_summary,
             caption = "candidate_microbe rows by method × direction (in DB)")
cat("Total candidate_microbe rows in DB:", n_total, "\n")
```

## Persist

```{r}
#| label: persist
write_csv(df_test,           path_target("wilcoxon_full_results.csv"))
write_csv(df_maaslin,        path_target("maaslin2_full_results.csv"))
write_csv(df_ancom_raw,      path_target("ancombc_full_results.csv"))
write_csv(df_deseq,          path_target("deseq2_full_results.csv"))
write_csv(df_all_methods,    path_target("candidate_microbe.csv"))
write_csv(df_agree,          path_target("method_agreement.csv"))
write_csv(df_consensus,      path_target("consensus_species.csv"))
```

## Cross-cohort meta-analysis (050-r2)

Combine per-cohort Wilcoxon statistics across the six Geng 2024 sub-cohorts via
Stouffer's Z (sqrt-n weights, sign tracking on log2FC) and a DerSimonian–Laird
random-effects pool of rank-biserial correlations. Outputs two new
`candidate_microbe` method tags: `meta_stouffer` and `meta_random`, with
`n_cohorts >= 2`.

```{r}
#| label: meta-per-cohort-wilcoxon
con <- load_db()
df_sample_all <- read_table_db(con, "sample") |>
  dplyr::filter(grepl("^geng2024_", source_id),
                disease_status %in% c("periodontitis", "healthy"))
df_profile_all <- read_table_db(con, "taxon_profile") |>
  dplyr::filter(sample_id %in% df_sample_all$sample_id,
                taxon_kind == "bacterium")
close_db(con)

cohorts_meta <- sort(unique(df_sample_all$source_id))
cat("Meta-analysis cohorts:", length(cohorts_meta), "\n")

per_cohort_wilcox <- function(source_id_x) {
  df_s <- df_sample_all |> dplyr::filter(source_id == source_id_x)
  df_p <- df_profile_all |> dplyr::filter(sample_id %in% df_s$sample_id)
  df_w <- df_p |>
    dplyr::select(sample_id, taxon_id, taxon_name, abundance) |>
    tidyr::pivot_wider(id_cols = c(taxon_id, taxon_name),
                       names_from = sample_id, values_from = abundance,
                       values_fill = 0)
  ids_case_x <- df_s$sample_id[df_s$disease_status == "periodontitis"]
  ids_ctrl_x <- df_s$sample_id[df_s$disease_status == "healthy"]
  if (length(ids_case_x) < 5 || length(ids_ctrl_x) < 5) return(NULL)
  mat_x <- as.matrix(df_w[, c(ids_case_x, ids_ctrl_x)])
  rownames(mat_x) <- df_w$taxon_id
  prev_x <- rowMeans(mat_x > 0)
  keep_x <- prev_x >= 0.10
  if (sum(keep_x) == 0) return(NULL)
  mat_case <- mat_x[keep_x, ids_case_x, drop = FALSE]
  mat_ctrl <- mat_x[keep_x, ids_ctrl_x, drop = FALSE]
  n1 <- length(ids_case_x); n2 <- length(ids_ctrl_x); n_total <- n1 + n2
  taxa_x <- rownames(mat_case)
  out <- purrr::map_dfr(seq_along(taxa_x), function(i) {
    a <- mat_case[i, ]; b <- mat_ctrl[i, ]
    if (length(unique(c(a, b))) < 2) {
      return(tibble::tibble(taxon_id = taxa_x[i],
                            p_value = NA_real_, U = NA_real_,
                            r = NA_real_, var_r = NA_real_,
                            log2fc = 0, mean_case = mean(a),
                            mean_ctrl = mean(b)))
    }
    wt <- suppressWarnings(stats::wilcox.test(a, b, exact = FALSE))
    U <- unname(wt$statistic)
    r <- 2 * U / (n1 * n2) - 1        # positive r = case-enriched
    var_r <- (1 - r^2)^2 / max(n_total - 1, 1)
    if (var_r < 1e-8) var_r <- 1e-8
    tibble::tibble(taxon_id = taxa_x[i],
                   p_value = wt$p.value, U = U,
                   r = r, var_r = var_r,
                   log2fc = log2((mean(a) + 1e-6) / (mean(b) + 1e-6)),
                   mean_case = mean(a), mean_ctrl = mean(b))
  })
  out |>
    dplyr::left_join(df_w |> dplyr::select(taxon_id, taxon_name),
                     by = "taxon_id") |>
    dplyr::mutate(source_id = source_id_x, n_case = n1, n_ctrl = n2,
                  n_total = n_total) |>
    dplyr::filter(!is.na(p_value))
}

df_per_cohort <- purrr::map_dfr(cohorts_meta, per_cohort_wilcox)

df_per_cohort |>
  dplyr::count(source_id, name = "n_species_tested") |>
  knitr::kable(caption = "Per-cohort Wilcoxon: species tested (prevalence >= 10%)")
```

```{r}
#| label: meta-combine
df_per_cohort_clean <- df_per_cohort |>
  dplyr::filter(is.finite(p_value), p_value > 0, p_value < 1,
                is.finite(r), is.finite(var_r))

df_grouped <- df_per_cohort_clean |>
  dplyr::group_by(taxon_id) |>
  dplyr::summarise(
    n_cohorts  = dplyr::n_distinct(source_id),
    taxon_name = dplyr::first(stats::na.omit(taxon_name)),
    .groups = "drop"
  ) |>
  dplyr::filter(n_cohorts >= 2)

cat("Species testable in >= 2 cohorts:", nrow(df_grouped), "\n")

stouffer_one <- function(df_one) {
  pv  <- df_one$p_value
  fc  <- df_one$log2fc
  n   <- df_one$n_total
  z_two <- stats::qnorm(1 - pv / 2)
  z_two[!is.finite(z_two)] <- 0
  z_signed <- z_two * ifelse(fc >= 0, 1, -1)
  w <- sqrt(n)
  z_pool <- sum(w * z_signed) / sqrt(sum(w^2))
  p_pool <- 2 * stats::pnorm(-abs(z_pool))
  tibble::tibble(z_pool = z_pool, p_value = p_pool,
                 effect_size = mean(fc),
                 direction   = ifelse(z_pool >= 0, "disease_enriched",
                                                   "health_enriched"))
}

df_meta_stouffer <- df_per_cohort_clean |>
  dplyr::semi_join(df_grouped, by = "taxon_id") |>
  dplyr::group_by(taxon_id) |>
  dplyr::group_modify(~ stouffer_one(.x)) |>
  dplyr::ungroup() |>
  dplyr::left_join(df_grouped, by = "taxon_id") |>
  dplyr::mutate(q_value = stats::p.adjust(p_value, method = "BH"))

cat("Stouffer meta: ", nrow(df_meta_stouffer),
    " species | q<0.1: ", sum(df_meta_stouffer$q_value < 0.1, na.rm = TRUE),
    "\n", sep = "")

has_metafor <- requireNamespace("metafor", quietly = TRUE)
random_one <- function(df_one) {
  yi <- df_one$r; vi <- df_one$var_r
  if (has_metafor) {
    fit <- try(metafor::rma(yi = yi, vi = vi, method = "DL"), silent = TRUE)
    if (!inherits(fit, "try-error")) {
      return(tibble::tibble(effect_size = unname(fit$b[1, 1]),
                            se          = fit$se,
                            p_value     = fit$pval,
                            tau2        = fit$tau2,
                            I2          = fit$I2))
    }
  }
  # DL fallback: IVW
  w <- 1 / vi
  est <- sum(w * yi) / sum(w)
  se  <- sqrt(1 / sum(w))
  z   <- est / se
  tibble::tibble(effect_size = est, se = se,
                 p_value     = 2 * stats::pnorm(-abs(z)),
                 tau2 = NA_real_, I2 = NA_real_)
}

df_meta_random <- df_per_cohort_clean |>
  dplyr::semi_join(df_grouped, by = "taxon_id") |>
  dplyr::group_by(taxon_id) |>
  dplyr::group_modify(~ random_one(.x)) |>
  dplyr::ungroup() |>
  dplyr::left_join(df_grouped, by = "taxon_id") |>
  dplyr::mutate(q_value   = stats::p.adjust(p_value, method = "BH"),
                direction = ifelse(effect_size >= 0, "disease_enriched",
                                                     "health_enriched"))

cat("Random-effects meta (",
    ifelse(has_metafor, "DL via metafor", "DL fallback: IVW"),
    "): ", nrow(df_meta_random),
    " species | q<0.1: ", sum(df_meta_random$q_value < 0.1, na.rm = TRUE),
    "\n", sep = "")
```

```{r}
#| label: meta-write-db
df_cand_meta_stouffer <- df_meta_stouffer |>
  dplyr::transmute(taxon_id, taxon_name, direction,
                   effect_size, p_value, q_value, n_cohorts) |>
  dplyr::mutate(
    phenotype      = "periodontitis",
    method         = "meta_stouffer",
    schema_version = ver,
    candidate_id   = paste0(
      "CAND_",
      vapply(paste(taxon_id, phenotype, method, sep = "|"),
             function(s) substr(digest::digest(s, algo = "xxhash64"), 1, 12),
             character(1))
    )
  ) |>
  dplyr::select(candidate_id, taxon_id, taxon_name, phenotype, direction,
                effect_size, p_value, q_value, n_cohorts, method,
                schema_version)

df_cand_meta_random <- df_meta_random |>
  dplyr::transmute(taxon_id, taxon_name, direction,
                   effect_size, p_value, q_value, n_cohorts) |>
  dplyr::mutate(
    phenotype      = "periodontitis",
    method         = "meta_random",
    schema_version = ver,
    candidate_id   = paste0(
      "CAND_",
      vapply(paste(taxon_id, phenotype, method, sep = "|"),
             function(s) substr(digest::digest(s, algo = "xxhash64"), 1, 12),
             character(1))
    )
  ) |>
  dplyr::select(candidate_id, taxon_id, taxon_name, phenotype, direction,
                effect_size, p_value, q_value, n_cohorts, method,
                schema_version)

df_cand_meta <- dplyr::bind_rows(df_cand_meta_stouffer, df_cand_meta_random)
stopifnot(!any(duplicated(df_cand_meta$candidate_id)))

con <- load_db()
DBI::dbExecute(con, "DELETE FROM candidate_microbe WHERE method LIKE 'meta_%'")
DBI::dbWriteTable(con, "candidate_microbe", df_cand_meta,
                  append = TRUE, row.names = FALSE)
df_meta_summary <- DBI::dbGetQuery(con,
  "SELECT method, COUNT(*) n, MIN(n_cohorts) min_c, MAX(n_cohorts) max_c
     FROM candidate_microbe GROUP BY method ORDER BY method")
close_db(con)

knitr::kable(df_meta_summary,
             caption = "candidate_microbe by method after meta merge")

df_meta_stouffer |>
  dplyr::arrange(q_value) |>
  head(10) |>
  dplyr::select(taxon_id, taxon_name, n_cohorts, direction,
                z_pool, p_value, q_value) |>
  knitr::kable(digits = 4,
               caption = "Top 10 species by Stouffer meta q_value")

df_meta_random |>
  dplyr::arrange(q_value) |>
  head(10) |>
  dplyr::select(taxon_id, taxon_name, n_cohorts, direction,
                effect_size, se, p_value, q_value) |>
  knitr::kable(digits = 4,
               caption = "Top 10 species by random-effects meta q_value")

write_csv(df_per_cohort,    path_target("meta_per_cohort_wilcoxon.csv"))
write_csv(df_meta_stouffer, path_target("meta_stouffer.csv"))
write_csv(df_meta_random,   path_target("meta_random.csv"))
```

## Files written

```{r}
#| label: list-files-target
knitr::kable(qproj::proj_dir_info(path_target(), tz = "CET"))
```
