---
title: "060-m1a-vsc-association"
date: today
params:
  name: "060-m1a-vsc-association"
---

**Updated: `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'CET')` CET.**

::: {.callout-note}
## Deferred to ripple

VSC (volatile sulfur compound) and clinical-marker association is **not in
the MVP scope**. The PRJDB11203 sub-cohort used in 040/050 is a
periodontitis case-control without VSC measurements, so there is nothing to
correlate at this point in the pipeline.

**Ripple plan** (`060-r1`):

- Identify cohorts with paired VSC / pocket depth / inflammation measurements
  (the halitosis–VSC reviews referenced in `analyses/background.md` give a
  starting list).
- Register them in `data_source` (`020-r1`) and ingest in `040-r1`.
- Run quantitative association: Spearman rank-correlation against VSC
  concentration; partial correlations adjusting for age/sex; mediation
  analysis (microbe → VSC → halitosis severity).
- Write candidates with `phenotype = "halitosis"` into `candidate_microbe`.
:::

```{r}
#| label: params
#| eval: !expr interactive()
#| include: false
params <- list(name = "060-m1a-vsc-association")
```

```{r}
#| label: setup
#| include: false
#| message: false
#| warning: false
#| cache: false
wd <- "analyses"
if (basename(getwd()) != wd) {
    setwd(here::here(wd))
}
here::i_am(paste0(params$name, ".qmd"), uuid = "9c5e3d65-bda3-4d29-87f4-0bcaa1aa1ddb")
qproj::proj_create_dir_target(params$name, clean = FALSE)
path_target <- qproj::proj_path_target(params$name)
```

## Status

```{r}
#| label: status
tibble::tibble(
  module          = "060-m1a-vsc-association",
  status          = "deferred-to-ripple",
  ripple_targets  = "060-r1 (Spearman + partial correlation against VSC / pocket depth)",
  blocker_reason  = "MVP cohort has no VSC measurements",
  next_action     = "Register a halitosis cohort with VSC in 020-r1, then run this module"
) |> knitr::kable(caption = "Module status")
```

## Files written

```{r}
#| label: list-files-target
knitr::kable(qproj::proj_dir_info(path_target(), tz = "CET"))
```
