---
title: "070-m2a-virome-processing"
date: today
params:
  name: "070-m2a-virome-processing"
bibliography: references.bib
---

**Updated: `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'CET')` CET.**

Loads the Yahara et al. 2021 long-read oral virome (PromethION saliva
sequencing, 4 samples, VIRSorter-derived contigs). Populates four `sample`
rows (saliva, healthy donor) and `taxon_profile` rows (`taxon_kind = "virus"`,
`abundance_kind = "presence"` since per-sample read counts are sparse and
inconsistent across the four sheets).

VIRSorter category 3 and category 6 contigs are dropped per the paper's
convention (low-confidence phage/prophage calls).

**Ripple ladder**:

- `070-r1`: re-process the raw PromethION fastq via ViWrap on SLURM, run CheckV
  for completeness/quality flags, dereplicate at 95% ANI, regenerate vOTU IDs.
- `070-r2`: AMG annotation (DRAM-v) and lifestyle (lytic vs temperate) calls.

```{r}
#| label: params
#| eval: !expr interactive()
#| include: false
params <- list(name = "070-m2a-virome-processing")
```

```{r}
#| label: setup
#| include: false
#| message: false
#| warning: false
#| cache: false
wd <- "analyses"
if (basename(getwd()) != wd) {
    setwd(here::here(wd))
}
here::i_am(paste0(params$name, ".qmd"), uuid = "8df56fa1-e1f5-484f-b4dd-4b4c9b56b4cf")
qproj::proj_create_dir_target(params$name, clean = FALSE)

path_target <- qproj::proj_path_target(params$name)
```

```{r}
#| label: packages
#| message: false
#| warning: false
#| cache: false
suppressPackageStartupMessages({
  library(here)
  library(conflicted)
  library(tidyverse)
  library(readxl)
  library(jsonlite)
  library(digest)
  library(DBI)
  devtools::load_all()
})
conflicted::conflicts_prefer(
  dplyr::filter,
  dplyr::lag,
  dplyr::first,
  dplyr::last,
  dplyr::between,
  .quiet = TRUE
)
```

## Load Yahara MOESM7 (Supp Data 4: VIRSorter contigs)

```{r}
#| label: load
#| cache: true
path_xlsx <- here::here("data", "00-raw", "d020-data-source-registry",
                        "yahara2021", "MOESM7.xlsx")
stopifnot(file.exists(path_xlsx))

sheets <- readxl::excel_sheets(path_xlsx)
df_phage_raw <- purrr::map_dfr(sheets, function(s) {
  d <- readxl::read_excel(path_xlsx, sheet = s, skip = 2,
                          .name_repair = "universal_quiet")
  d$sample_label <- s  # Sample1 .. Sample4
  d
})
cat("Total contigs across 4 sheets:", nrow(df_phage_raw), "\n")
print(dplyr::count(df_phage_raw, sample_label, name = "n_contigs"))
```

## Filter and shape

```{r}
#| label: shape
df_phage <- df_phage_raw |>
  dplyr::filter(!is.na(phageID),
                category..VirSorter. %in% c(1, 2, 4, 5)) |>
  dplyr::mutate(
    sample_id      = paste0("SAMP_yahara2021_", sample_label),
    vOTU_index     = sprintf("%05d", dplyr::row_number()),
    taxon_id       = paste0("vOTU_yahara2021_", vOTU_index),
    taxon_kind     = "virus",
    taxon_name     = phageID,
    abundance      = 1.0,
    abundance_kind = "presence"
  )
cat("After category filter:", nrow(df_phage),
    "| samples:", dplyr::n_distinct(df_phage$sample_id),
    "| unique vOTU IDs:", dplyr::n_distinct(df_phage$taxon_id), "\n")
```

## Build `sample` rows for the four Yahara saliva samples

```{r}
#| label: build-sample
df_sample_yahara <- tibble::tibble(
  sample_id      = paste0("SAMP_yahara2021_", sheets),
  source_id      = "yahara2021",
  subject_id     = "yahara2021_donor",  # one donor across the four conditions
  disease_status = "healthy",
  body_site      = "saliva",
  seq_type       = "virome",
  meta_json      = jsonlite::toJSON(
    list(platform = "PromethION + HiSeq",
         note     = "single donor x 4 conditions"),
    auto_unbox = TRUE
  ) |> rep(length(sheets))
)
df_sample_yahara
```

## Build `taxon_profile` rows

```{r}
#| label: build-taxon-profile
df_taxon_profile_v <- df_phage |>
  dplyr::mutate(
    profile_id = paste0(
      "PROF_",
      vapply(paste0(sample_id, "|", taxon_id),
             function(s) substr(digest::digest(s, algo = "xxhash64"), 1, 12),
             character(1))
    )
  ) |>
  dplyr::select(profile_id, sample_id, taxon_id, taxon_kind, taxon_name,
                abundance, abundance_kind)
stopifnot(!any(duplicated(df_taxon_profile_v$profile_id)))
cat("virus taxon_profile rows:", nrow(df_taxon_profile_v), "\n")
```

## Stash phage metadata for 080 (host prediction)

The full phage table — including the CAT-assigned host taxonomy — is the
input to module 080. Persist it as a CSV that 080 will read.

```{r}
#| label: stash-phage-meta
df_phage_for_080 <- df_phage |>
  dplyr::transmute(
    sample_id,
    vOTU_id              = taxon_id,
    phageID,
    phageLen,
    linear_or_circular   = linear.or.circular,
    integrase            = Integrase,
    virsorter_category   = category..VirSorter.,
    cat_taxonomy         = taxonomy.assigned.to.the.contig.by.CAT,
    img_vr_cluster       = clustered.with.IMG.VR.v2.0,
    assigned_family      = assigned.family
  )
write_csv(df_phage_for_080, path_target("phage_meta_for_080.csv"))
```

## Write to the contract layer (append, since 040 already wrote bacterial samples + profiles)

```{r}
#| label: write-db
con <- load_db()
assert_schema(con)

# Append the 4 Yahara samples (don't truncate the 42 Geng samples 040 wrote)
DBI::dbExecute(con,
  "DELETE FROM sample WHERE source_id = 'yahara2021'")
DBI::dbWriteTable(con, "sample", df_sample_yahara, append = TRUE,
                  row.names = FALSE)

DBI::dbExecute(con,
  "DELETE FROM taxon_profile WHERE taxon_kind = 'virus'")
DBI::dbWriteTable(con, "taxon_profile", df_taxon_profile_v, append = TRUE,
                  row.names = FALSE)

df_check <- tibble::tibble(
  table = c("sample", "taxon_profile_bacteria", "taxon_profile_virus"),
  rows  = c(
    DBI::dbGetQuery(con, "SELECT COUNT(*) AS n FROM sample")$n,
    DBI::dbGetQuery(con, "SELECT COUNT(*) AS n FROM taxon_profile WHERE taxon_kind='bacterium'")$n,
    DBI::dbGetQuery(con, "SELECT COUNT(*) AS n FROM taxon_profile WHERE taxon_kind='virus'")$n
  )
)
close_db(con)
knitr::kable(df_check, caption = "Contract tables after 070 write")
```

## Files written

```{r}
#| label: list-files-target
knitr::kable(qproj::proj_dir_info(path_target(), tz = "CET"))
```
