---
title: "080-m2a-host-prediction"
date: today
params:
  name: "080-m2a-host-prediction"
bibliography: references.bib
---

**Updated: `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'CET')` CET.**

Parses the CAT taxonomy column from the Yahara 2021 viral contig table
(stashed by 070) into structured `phage_host_link` rows. CAT assigns each
contig a single best-rank taxonomy string like `Granulicatella (genus): 0.53`.
The numeric tail is the CAT confidence score (0..1) which we propagate as
`confidence`. `evidence_kind = "published_prediction"` since these calls
come from the paper as-is.

We keep only **genus-level or finer** assignments — superkingdom / phylum /
"no rank" rows are not actionable for host-microbe matching.

**Ripple ladder**:

- `080-r1`: ensemble — re-run host prediction with iPHoP, RaFAH, spacepharer
  CRISPR matches against eHOMD, BLAST against IMG/VR, and combine via
  precision-recall–calibrated voting.
- `080-r2`: deep-learning predictors (vHULK, PHIST) for the long contigs.

```{r}
#| label: params
#| eval: !expr interactive()
#| include: false
params <- list(name = "080-m2a-host-prediction")
```

```{r}
#| label: setup
#| include: false
#| message: false
#| warning: false
#| cache: false
wd <- "analyses"
if (basename(getwd()) != wd) {
    setwd(here::here(wd))
}
here::i_am(paste0(params$name, ".qmd"), uuid = "d8f9e132-2c2d-4bd9-9a3e-c1bda3e8c8e4")
qproj::proj_create_dir_target(params$name, clean = FALSE)

path_target <- qproj::proj_path_target(params$name)
```

```{r}
#| label: packages
#| message: false
#| warning: false
#| cache: false
suppressPackageStartupMessages({
  library(here)
  library(conflicted)
  library(tidyverse)
  library(jsonlite)
  library(digest)
  library(DBI)
  devtools::load_all()
})
conflicted::conflicts_prefer(
  dplyr::filter,
  dplyr::lag,
  dplyr::first,
  dplyr::last,
  dplyr::between,
  .quiet = TRUE
)
```

## Pull the stashed Yahara contig metadata

```{r}
#| label: pull
path_in <- here::here("data", "070-m2a-virome-processing",
                      "phage_meta_for_080.csv")
stopifnot(file.exists(path_in))
df_phage_meta <- readr::read_csv(path_in, show_col_types = FALSE)
cat("Input rows:", nrow(df_phage_meta), "\n")
```

## Parse the CAT taxonomy string

```{r}
#| label: parse-cat
slugify <- function(x) {
  x |>
    iconv(to = "ASCII//TRANSLIT") |>
    gsub("[^A-Za-z0-9]+", "_", x = _) |>
    gsub("_+$", "", x = _) |>
    gsub("^_+", "", x = _)
}

# CAT format: "<name> (<rank>): <conf>"
re_cat <- "^(.*?)\\s*\\(([^)]+)\\):\\s*([0-9.]+)$"

df_parsed <- df_phage_meta |>
  dplyr::filter(!is.na(cat_taxonomy)) |>
  dplyr::mutate(
    host_name        = sub(re_cat, "\\1", cat_taxonomy),
    host_rank        = sub(re_cat, "\\2", cat_taxonomy),
    confidence       = suppressWarnings(as.numeric(sub(re_cat, "\\3", cat_taxonomy)))
  ) |>
  dplyr::filter(!is.na(confidence))

cat("After parse:", nrow(df_parsed), "\n")
cat("Rank distribution:\n")
print(dplyr::count(df_parsed, host_rank, sort = TRUE))
```

## Restrict to actionable ranks (genus / species / etc.)

```{r}
#| label: filter-ranks
actionable_ranks <- c("genus", "species", "subspecies", "strain")
df_actionable <- df_parsed |>
  dplyr::filter(host_rank %in% actionable_ranks) |>
  dplyr::mutate(
    host_taxon_id = paste0("STR_", slugify(host_name))
  )
cat("Actionable host-link rows:", nrow(df_actionable), "\n")
print(dplyr::count(df_actionable, host_rank))
```

## Build `phage_host_link` rows

```{r}
#| label: build-links
ver <- schema_version()

df_phage_host_link <- df_actionable |>
  dplyr::transmute(
    vOTU_id        = vOTU_id,
    host_taxon_id,
    evidence_kind  = "published_prediction",
    confidence,
    support_json   = purrr::pmap_chr(
      list(host_name, host_rank, integrase, virsorter_category, img_vr_cluster),
      function(host_name, host_rank, integrase, virsorter_category, img_vr_cluster) {
        jsonlite::toJSON(
          list(host_name        = host_name,
               host_rank        = host_rank,
               integrase        = integrase,
               virsorter_cat    = virsorter_category,
               img_vr_cluster   = img_vr_cluster,
               method           = "CAT (Yahara 2021)"),
          auto_unbox = TRUE, na = "null"
        )
      }
    ),
    schema_version = ver
  ) |>
  dplyr::mutate(
    link_id = paste0(
      "LINK_",
      vapply(paste(vOTU_id, host_taxon_id, evidence_kind, sep = "|"),
             function(s) substr(digest::digest(s, algo = "xxhash64"), 1, 12),
             character(1))
    )
  ) |>
  dplyr::distinct(link_id, .keep_all = TRUE) |>
  dplyr::select(link_id, vOTU_id, host_taxon_id, evidence_kind,
                confidence, support_json, schema_version)

cat("phage_host_link rows:", nrow(df_phage_host_link),
    "| unique vOTUs:", dplyr::n_distinct(df_phage_host_link$vOTU_id),
    "| unique hosts:", dplyr::n_distinct(df_phage_host_link$host_taxon_id), "\n")
df_phage_host_link |>
  dplyr::count(host_taxon_id, sort = TRUE) |>
  head(15) |>
  knitr::kable(caption = "Top 15 predicted hosts by phage-link count")
```

## Write to the contract layer

```{r}
#| label: write-db
con <- load_db()
assert_schema(con)
write_table_db(con, "phage_host_link", df_phage_host_link, append = FALSE)
n <- DBI::dbGetQuery(con, "SELECT COUNT(*) AS n FROM phage_host_link")$n
close_db(con)
cat("phage_host_link rows in DB:", n, "\n")
```

## Persist

```{r}
#| label: persist
write_csv(df_phage_host_link, path_target("phage_host_link.csv"))
```

## Files written

```{r}
#| label: list-files-target
knitr::kable(qproj::proj_dir_info(path_target(), tz = "CET"))
```
