---
title: "090-m2a-mapping"
date: today
params:
  name: "090-m2a-mapping"
bibliography: references.bib
---

**Updated: `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'CET')` CET.**

Joins `candidate_microbe` (M1a output) with `phage_host_link` (M2a output)
to surface the phages whose predicted hosts overlap with periodontitis
candidates. The MVP uses two join strategies:

1. **Strict** — `candidate_microbe.taxon_id == phage_host_link.host_taxon_id`.
   Likely zero hits in this MVP because Geng's bacteriome is at species
   level (e.g. `STR_Streptococcus_salivarius`) while Yahara's CAT
   predictions are at genus level (e.g. `STR_Streptococcus`).
2. **Genus prefix** — candidate species belongs to the predicted host
   genus (`candidate.taxon_id LIKE host_taxon_id || '_%'`). This is the
   primary mapping for MVP downstream use.

`100-knowledge-integration` consumes the genus-prefix mapping to assemble
`evidence_packet`. This module's outputs are exploratory artefacts (CSVs +
plots), not contract-layer writes.

**Ripple ladder**:

- `090-r1`: replace genus prefix with eHOMD-aware fuzzy taxonomy (handle
  synonyms, basonyms, misspellings, Candidatus naming).
- `090-r2`: bipartite-graph visualizations (heatmap, network); centrality
  metrics; partitioning of phage–host modules.

```{r}
#| label: params
#| eval: !expr interactive()
#| include: false
params <- list(name = "090-m2a-mapping")
```

```{r}
#| label: setup
#| include: false
#| message: false
#| warning: false
#| cache: false
wd <- "analyses"
if (basename(getwd()) != wd) {
    setwd(here::here(wd))
}
here::i_am(paste0(params$name, ".qmd"), uuid = "75d27a42-be91-466a-a4ed-1f4c2ddf9e30")
qproj::proj_create_dir_target(params$name, clean = FALSE)

path_target <- qproj::proj_path_target(params$name)
```

```{r}
#| label: packages
#| message: false
#| warning: false
#| cache: false
suppressPackageStartupMessages({
  library(here)
  library(conflicted)
  library(tidyverse)
  library(DBI)
  devtools::load_all()
})
conflicted::conflicts_prefer(
  dplyr::filter,
  dplyr::lag,
  dplyr::first,
  dplyr::last,
  dplyr::between,
  .quiet = TRUE
)
```

## Pull contract tables

```{r}
#| label: pull
con <- load_db()
df_candidate <- read_table_db(con, "candidate_microbe")
df_link      <- read_table_db(con, "phage_host_link")
close_db(con)
cat("candidate_microbe:", nrow(df_candidate),
    " | phage_host_link:", nrow(df_link), "\n")
```

## Strategy 1: strict equality

```{r}
#| label: strict
df_strict <- df_candidate |>
  dplyr::inner_join(df_link, by = c("taxon_id" = "host_taxon_id")) |>
  dplyr::select(candidate_id, taxon_id, taxon_name, phenotype, direction,
                effect_size, q_value,
                link_id, vOTU_id, evidence_kind, confidence)
cat("Strict-equality hits:", nrow(df_strict), "\n")
```

## Strategy 2: genus-prefix join (MVP primary)

```{r}
#| label: prefix
df_prefix <- df_candidate |>
  dplyr::inner_join(df_link,
                    by = character(),  # cross join
                    relationship = "many-to-many") |>
  dplyr::filter(stringr::str_starts(taxon_id,
                                    paste0(host_taxon_id, "_"))) |>
  dplyr::select(candidate_id, taxon_id, taxon_name, phenotype, direction,
                effect_size, q_value,
                link_id, vOTU_id, host_taxon_id, evidence_kind, confidence)
cat("Genus-prefix hits:",
    nrow(df_prefix),
    " | unique candidates with phage hits:",
    dplyr::n_distinct(df_prefix$candidate_id),
    " | unique phages:",
    dplyr::n_distinct(df_prefix$vOTU_id), "\n")
```

## Per-genus summary

```{r}
#| label: per-genus
df_genus_summary <- df_prefix |>
  dplyr::group_by(host_taxon_id) |>
  dplyr::summarise(
    n_candidates = dplyr::n_distinct(candidate_id),
    n_phages     = dplyr::n_distinct(vOTU_id),
    direction_breakdown = paste(
      paste0(unique(direction), "(",
             vapply(unique(direction), function(d) dplyr::n_distinct(candidate_id[direction == d]), integer(1)),
             ")"),
      collapse = ", "
    ),
    mean_confidence = mean(confidence, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::arrange(dplyr::desc(n_candidates), dplyr::desc(n_phages))
knitr::kable(df_genus_summary, caption = "Phage–host genus mapping summary")
```

## Persist mapping artefacts

```{r}
#| label: persist
write_csv(df_strict, path_target("mapping_strict.csv"))
write_csv(df_prefix, path_target("mapping_prefix.csv"))
write_csv(df_genus_summary, path_target("mapping_genus_summary.csv"))
```

## Quick visualization

```{r}
#| label: fig-mapping-genus
#| fig-width: 7
#| fig-height: 4
df_plot <- df_genus_summary |>
  dplyr::slice_max(n_phages, n = 12) |>
  dplyr::mutate(host = sub("^STR_", "", host_taxon_id))
p <- ggplot(df_plot,
            aes(x = reorder(host, n_phages), y = n_phages)) +
  geom_col(aes(fill = n_candidates)) +
  coord_flip() +
  scale_fill_viridis_c(name = "Candidate species") +
  labs(x = NULL, y = "Phages predicted to infect this genus",
       title = "M2a phage hits per M1a-implicated host genus") +
  theme_minimal(base_size = 11)
ggsave(path_target("fig_genus_mapping.png"),
       p, width = 7, height = 4, dpi = 150)
print(p)
```

## Files written

```{r}
#| label: list-files-target
knitr::kable(qproj::proj_dir_info(path_target(), tz = "CET"))
```
