---
title: "Oral probiotic phage knowledge layer — MVP data report"
subtitle: "Stub: numbers and figure links only; prose deferred to ripple."
date: today
params:
  name: "990-manuscript"
bibliography: references.bib
---

::: {.callout-note}
## MVP scope

This is the manuscript **stub**. It references the schema, summarises the
contract-table populations, and links to the figures from 110. No
introduction / discussion is written yet — that is `990-r1`. When ripples
land enough cohorts (`020-r1`, `040-r1`) and methods (`050-r1`, `080-r1`),
the prose will fold around this stub.
:::

```{r}
#| label: params
#| eval: !expr interactive()
#| include: false
params <- list(name = "990-manuscript")
```

```{r}
#| label: setup
#| include: false
#| message: false
#| warning: false
#| cache: false
wd <- "analyses"
if (basename(getwd()) != wd) {
    setwd(here::here(wd))
}
here::i_am(paste0(params$name, ".qmd"), uuid = "9c3df0cc-2706-456c-a93b-d92d52cd2c7c")
qproj::proj_create_dir_target(params$name, clean = FALSE)

path_target <- qproj::proj_path_target(params$name)
```

```{r}
#| label: packages
#| include: false
#| message: false
#| warning: false
#| cache: false
suppressPackageStartupMessages({
  library(here)
  library(tidyverse)
  library(yaml)
  library(DBI)
  devtools::load_all()
})
```

# Data layer

The contract layer defines five core objects plus a data-source registry
([Sample, TaxonProfile, CandidateMicrobe, PhageHostLink, EvidencePacket]).
Schema version: `r schema_version()`; canonical YAML lives at
`analyses/data/010-schema-design/schema-v0.1.yml`.

```{r}
#| label: schema-summary
#| echo: false
schema <- yaml::read_yaml(schema_path())
df_schema_summary <- tibble::tibble(
  table     = names(schema$tables),
  pk        = vapply(schema$tables, function(t) t$pk %||% "", character(1)),
  n_columns = vapply(schema$tables, function(t) length(t$columns), integer(1))
)
knitr::kable(df_schema_summary, caption = "Schema v0.1 — six contract tables")
```

# Sources used in the MVP

`r if(file.exists(here::here("data","020-data-source-registry","data_source.csv"))) "" else "(data_source.csv not found)"`

```{r}
#| label: sources
#| echo: false
con <- load_db()
df_source <- read_table_db(con, "data_source")
df_summary <- tibble::tibble(
  table = c("data_source","sample","taxon_profile","candidate_microbe",
            "phage_host_link","evidence_packet"),
  rows  = c(
    DBI::dbGetQuery(con, "SELECT COUNT(*) AS n FROM data_source")$n,
    DBI::dbGetQuery(con, "SELECT COUNT(*) AS n FROM sample")$n,
    DBI::dbGetQuery(con, "SELECT COUNT(*) AS n FROM taxon_profile")$n,
    DBI::dbGetQuery(con, "SELECT COUNT(*) AS n FROM candidate_microbe")$n,
    DBI::dbGetQuery(con, "SELECT COUNT(*) AS n FROM phage_host_link")$n,
    DBI::dbGetQuery(con, "SELECT COUNT(*) AS n FROM evidence_packet")$n
  )
)
close_db(con)
knitr::kable(df_source[, c("source_id","citation_key","doi","source_kind","n_samples")],
             caption = "Registered MVP data sources")
knitr::kable(df_summary, caption = "Contract-layer populations after MVP")
```

The bacteriome layer comes from the Geng et al. 2024 universal periodontitis
signature [@2024-Universal_Geng], specifically the PRJDB11203 saliva
sub-cohort. The virome layer comes from the Yahara et al. 2021 long-read
PromethION oral phageome [@2021-Longread_Yahara].

# MVP figures

- `analyses/data/090-m2a-mapping/fig_genus_mapping.png` — phages per
  M1a-implicated host genus.
- `analyses/data/110-evaluation/fig1_samples.png` — sample × source breakdown.
- `analyses/data/110-evaluation/fig2_volcano.png` — M1a Wilcoxon volcano.
- `analyses/data/110-evaluation/fig3_phage_coverage.png` — phage-coverage of
  M1a candidates.

# Methods (skeleton)

- M1a differential: Wilcoxon rank-sum on TSS relative abundances, BH FDR;
  filter species with prevalence ≥ 10%. See
  `analyses/050-m1a-differential-analysis.qmd`.
- M2a host prediction: CAT taxonomy of VIRSorter contigs from the published
  long-read assembly, restricted to genus-or-finer ranks. See
  `analyses/080-m2a-host-prediction.qmd`.
- Mapping: candidate species ↔ phage host via genus-prefix join. See
  `analyses/090-m2a-mapping.qmd`.

# What's deferred to ripple (manuscript-blocking items)

- multi-cohort meta-analysis (`050-r2`) → upgrade `n_cohorts ≥ 1` MVP
  candidates to robust signature.
- Real `compatibility_score` (`100-r1`) → currently `NA`.
- Cross-cohort virome (`020-r2` + `070-r1` on SLURM) → currently a single
  long-read paper.

# References {.unnumbered}

::: {#refs}
:::

## Files written

```{r}
#| label: list-files-target
knitr::kable(qproj::proj_dir_info(path_target(), tz = "CET"))
```
